### 2013-08-08 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding Newton-Richardson comparison

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Newton-Krylov
method. For each benchmark, we will analyze the iterative performance
of both the nonlinear Newton solver and the linear solver used to
compute the Newton correction vector at each step. Limiting the number
of nonlinear iterations required for solution is critical for
efficiency as nonlinear function evaluations, Jacobian generation, and
preconditioner construction are typically required at every Newton
iteration. In addition, timing results will also be discussed.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Rayleigh Number}& 
      \multicolumn{1}{c}{Newton-Krylov Iterations}&
      \multicolumn{1}{c}{FANM Iterations}\\
      \hline
      \sn{1}{3} & 5 & 5 \\
      \sn{1}{4} & 7 & 7 \\
      \sn{1}{5} & 9 & 10 \\
      \sn{1}{6} & 11 & 11 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Thermal convection cavity nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:convection_nonlinear_iter_comparison}
\end{table}

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solvers preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case where a constant forcing term of \sn{1}{-4} was kept, this
difference is most drastic with FANM requiring only 64\% of the linear
iterations required by the Newton-Krylov scheme with the linear solver
per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. If the performance of
the MCSA implementation used here were to be potentially optimized in
the future, this improved iterative performance could also mean
improved time to solution.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Rayleigh Number}& 
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}\\
      \hline
      \sn{1}{3} & 23 & 18 \\
      \sn{1}{4} & 23 & 17 \\
      \sn{1}{5} & 25 & 20 \\
      \sn{1}{6} & 39 & 25 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Thermal convection cavity total sum of linear
      iterations for all nonlinear iterations.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:convection_linear_iter_comparison}
\end{table}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/convection_ra1e6_iters.pdf}
  \end{center}
  \caption{\textbf{Thermal convection cavity linear solver iterations
      as a function of Newton iteration for the \sn{1}{6} Rayleigh
      number case.} \textit{MCSA shows better iterative performance
      than GMRES at all but the second nonlinear iteration. The
      forcing term was fixed at \sn{1}{-4} at all nonlinear
      iterations.}}
  \label{fig:ra1e6_linear_iters}
\end{figure}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{Newton-Krylov Iterations}&
      \multicolumn{1}{c}{FANM Iterations}\\
      \hline
      100 & 6 & 6 \\
      300 & 9 & 9 \\
      500 & 11 & 11 \\
      700 & 14 & 10 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Lid driven cavity nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:driven_nonlinear_iter_comparison}
\end{table}

@@

The difficulty of this benchmark problem as compared to the thermal
convection cavity problem is readily apparent in the linear iteration
counts as given by Table~\ref{tab:driven_linear_iter_comparison}. As
the opposite of what was observed for that benchmark, lid driven
cavity results show that the linear models were significantly more
ill-conditioned, requiring many more MCSA and GMRES iterations. For
all cases except the 700 Reynolds number case, many more MCSA
iterations were required to converge with the observed GMRES iteration
count 64\% of that observed for MCSA when looking at the 100 Reynolds
number case. For the 700 Reynolds number case, fewer MCSA iterations
were observed due to the fact that 4 fewer nonlinear iterations were
performed. The average number of linear iterations per nonlinear
iteration was still much larger for the FANM calculations.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}\\
      \hline
      100 & 27 & 42 \\
      300 & 35 & 52 \\
      500 & 41 & 56 \\
      700 & 21 & 14 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Lid driven cavity total sum of linear
      iterations for all nonlinear iterations.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:driven_linear_iter_comparison}
\end{table}

@@

The backward step problem exhibited different performance
characteristics than the other two benchmark problems, likely due to
the different geometry and boundary
conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
nonlinear iteration results for this benchmark. At low Reynolds
numbers, FANM converges in fewer nonlinear iterations than the
Newton-Krylov solver. However, as the problem becomes more difficult
to solver at higher rates of flow, the Newton-Krylov solver eventually
converges in fewer nonlinear iterations. We see the same behavior as
well in the linear solver iteration data given by
Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
numbers, FANM required significantly fewer linear solver iterations to
converge the problem while the ill-conditioned case at a Reynolds
number of 500 is observed to use many more MCSA iterations.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{Newton-Krylov Iterations}&
      \multicolumn{1}{c}{FANM Iterations}\\
      \hline
      200 & 10 & 9 \\
      300 & 15 & 14 \\
      400 & 10 & 10 \\
      500 & 19 & 20 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:step_nonlinear_iter_comparison}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}\\
      \hline
      200 & 24 & 13 \\
      300 & 23 & 17 \\
      400 & 18 & 12 \\
      500 & 30 & 52 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step total sum of linear
      iterations for all nonlinear iterations.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:step_linear_iter_comparison}
\end{table}

@@

>>>>

Using the benchmarks from the verification, we now compare the
performance of the FANM method against a conventional Newton-Krylov
(NK) method. For each benchmark, we will analyze the iterative
performance of both the nonlinear Newton solver and the linear solver
used to compute the Newton correction vector at each step. Limiting
the number of nonlinear iterations required for solution is critical
for efficiency as nonlinear function evaluations, Jacobian generation,
and preconditioner construction are typically required at every Newton
iteration. In addition, timing results will also be discussed. To
quantify the acceleration provided by the residual Monte Carlo
component of MCSA, each calculation was also performed using the same
Newton method but with a plain preconditioned Richardson iteration as
the linear solution technique to give a Newton-Richardson (NR)
scheme. For all cases the Richardson iteration was preconditioned
in the same way as GMRES and MCSA.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccc}\hline\hline
      \multicolumn{1}{c}{Rayleigh Number}& 
      \multicolumn{1}{c}{NK Iterations}&
      \multicolumn{1}{c}{FANM Iterations}&
      \multicolumn{1}{c}{NR Iterations}\\
      \hline
      \sn{1}{3} & 5 & 5 & 5\\
      \sn{1}{4} & 7 & 7 & 7\\
      \sn{1}{5} & 9 & 10 & 9\\
      \sn{1}{6} & 11 & 11 & 11\\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Thermal convection cavity nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:convection_nonlinear_iter_comparison}
\end{table}

@@

In addition to nonlinear iterative performance, performance of the
inner linear solves is also of interest. Here, we seek to maintain a
minimum number of total linear solver iterations over the course of
all nonlinear iterations in order to limit matrix-vector multiplies
and other costly
operations. Table~\ref{tab:convection_linear_iter_comparison} gives
the total number of GMRES iterations required to converge the problem
for the Newton-Krylov method and the total number of MCSA iterations
for the FANM method at each Rayleigh number used. For all cases, fewer
MCSA iterations were required than GMRES iterations to converge the
problem even with both solvers preconditioned with the same multigrid
package and the same parameters. For the \sn{1}{6} Rayleigh number
case where a constant forcing term of \sn{1}{-4} was kept, this
difference is most drastic with FANM requiring only 64\% of the linear
iterations required by the Newton-Krylov scheme with the linear solver
per nonlinear iteration given for this case in
Figure~\ref{fig:ra1e6_linear_iters}. Even the \sn{1}{5} Rayleigh
number case that required more Newton iterations required 5 fewer MCSA
iterations to converge than GMRES iterations. In all cases, MCSA
improved on the iterative performance of using only the Richardson
iteration through the residual Monte Carlo acceleration. If the performance of
the MCSA implementation used here were to be potentially optimized in
the future, this improved iterative performance could also mean
improved time to solution.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccc}\hline\hline
      \multicolumn{1}{c}{Rayleigh Number}&
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}&
      \multicolumn{1}{c}{Richardson Iterations}\\
      \hline
      \sn{1}{3} & 23 & 18 & 38 \\
      \sn{1}{4} & 23 & 17 & 34 \\
      \sn{1}{5} & 25 & 20 & 34 \\
      \sn{1}{6} & 39 & 25 & 48  \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Thermal convection cavity total sum of linear
      iterations for all nonlinear iterations.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:convection_linear_iter_comparison}
\end{table}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/convection_ra1e6_iters.pdf}
  \end{center}
  \caption{\textbf{Thermal convection cavity linear solver iterations
      as a function of Newton iteration for the \sn{1}{6} Rayleigh
      number case.} \textit{MCSA shows better iterative performance
      than GMRES at all but the second nonlinear iteration and MCSA
      always accelerated the Richardson iteration. The forcing term
      was fixed at \sn{1}{-4} at all nonlinear iterations.}}
  \label{fig:ra1e6_linear_iters}
\end{figure}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{NK Iterations}&
      \multicolumn{1}{c}{FANM Iterations}&
      \multicolumn{1}{c}{NR Iterations}\\
      \hline
      100 & 6 & 6 & 7\\
      300 & 9 & 9 & 9\\
      500 & 11 & 11 & 11\\
      700 & 14 & 10 & 12\\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Lid driven cavity nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:driven_nonlinear_iter_comparison}
\end{table}

@@

The difficulty of this benchmark problem as compared to the thermal
convection cavity problem is readily apparent in the linear iteration
counts as given by Table~\ref{tab:driven_linear_iter_comparison}. As
the opposite of what was observed for that benchmark, lid driven
cavity results show that the linear models were significantly more
ill-conditioned, requiring many more MCSA and GMRES iterations. For
all cases except the 700 Reynolds number case, many more MCSA
iterations were required to converge with the observed GMRES iteration
count 64\% of that observed for MCSA when looking at the 100 Reynolds
number case. For the 700 Reynolds number case, fewer MCSA iterations
were observed due to the fact that 4 fewer nonlinear iterations were
performed. The average number of linear iterations per nonlinear
iteration was still much larger for the FANM calculations. In all
cases, MCSA improved on the iterative performance of using only the
Richardson iteration through the residual Monte Carlo acceleration
even considering the cases with Reynolds numbers of 100 and 700 where
an extra Newton-Richardson iteration was required to converge the
nonlinear model.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}&
      \multicolumn{1}{c}{Richardson Iterations}\\
      \hline
      100 & 27 & 42 & 151\\
      300 & 35 & 52 & 133\\
      500 & 41 & 56 & 154\\
      700 & 21 & 14 & 32\\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Lid driven cavity total sum of linear iterations
      for all nonlinear iterations.} \textit{The results presented
      here were obtained from the benchmark verification
      calculations. MCSA always accelerated the Richardson
      iteration.}}
  \label{tab:driven_linear_iter_comparison}
\end{table}

@@

The backward step problem exhibited different performance
characteristics than the other two benchmark problems, likely due to
the different geometry and boundary
conditions. Table~\ref{tab:step_nonlinear_iter_comparison} gives the
nonlinear iteration results for this benchmark. At low Reynolds
numbers, FANM converges in fewer nonlinear iterations than the
Newton-Krylov solver. However, as the problem becomes more difficult
to solver at higher rates of flow, the Newton-Krylov solver eventually
converges in fewer nonlinear iterations. We see the same behavior as
well in the linear solver iteration data given by
Table~\ref{tab:step_linear_iter_comparison}. At the 3 lowest Reynolds
numbers, FANM required significantly fewer linear solver iterations to
converge the problem while the ill-conditioned case at a Reynolds
number of 500 is observed to use many more MCSA iterations. For
Reynolds numbers of both 200 and 500, FANM also reduced the number of
Newton steps need when compared to the Newton-Richardson scheme.

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{NK Iterations}&
      \multicolumn{1}{c}{FANM Iterations}&
      \multicolumn{1}{c}{NR Iterations}\\
      \hline
      200 & 10 & 9 & 10\\
      300 & 15 & 14 & 15\\
      400 & 10 & 10 & 10\\
      500 & 19 & 20 & 21\\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step nonlinear iteration
      results for the performance comparison.} \textit{The results
      presented here were obtained from the benchmark verification
      calculations.}}
  \label{tab:step_nonlinear_iter_comparison}
\end{table}

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cccc}\hline\hline
      \multicolumn{1}{c}{Reynolds Number}& 
      \multicolumn{1}{c}{GMRES Iterations}&
      \multicolumn{1}{c}{MCSA Iterations}&
      \multicolumn{1}{c}{Richardson Iterations}\\
      \hline
      200 & 24 & 13 & 21\\
      300 & 23 & 17 & 21\\
      400 & 18 & 12 & 14\\
      500 & 30 & 52 & 98\\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step total sum of linear iterations
      for all nonlinear iterations.} \textit{The results presented
      here were obtained from the benchmark verification
      calculations.MCSA always accelerated the Richardson iteration.}}
  \label{tab:step_linear_iter_comparison}
\end{table}

@@

We additionally see the effects of ill-conditioned Jacobians when
looking at the acceleration provided by MCSA over using a
preconditioned Richardson iteration. For the first 3 cases, the
acceleration was not as significant as for the other benchmark
problems. In this case, this means that the Richardson iteration and
its acceleration by a stochastic realization of a Richardson iteration
are not doing the bulk of the work in this problem. Rather, the
algebraic multigrid preconditioning is performing more work at each
iteration and therefore the acceleration is smaller. When the system
becomes even more ill-conditioned in the 500 Reynolds number case
the acceleration has more of an impact.
